Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-01 analysis/paper/hotspots-paper.Rmd
added chunk names to main document

@@

```{r}
if (!file.exists(here("analysis", "data", "derived_data", "expr.rds"))){
  download.file("https://datadryad.org/bitstream/handle/10255/dryad.166774/Attie_DO378_eQTL_viewer_v1.Rdata?sequence=2",
              destfile = here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata")
              )
  load(here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata"))
  saveRDS(dataset.islet.rnaseq$lod.peaks, here("analysis", "data", "derived_data", "lod_peaks.rds"))
  saveRDS(dataset.islet.rnaseq$annots, here("analysis", "data", "derived_data", "annots.rds"))
  saveRDS(dataset.islet.rnaseq$expr, here("analysis", "data", "derived_data", "expr.rds"))
  saveRDS(K, here("analysis", "data", "derived_data", "kinship.rds"))
  saveRDS(map, here("analysis", "data", "derived_data", "map.rds"))
  saveRDS(genoprobs, here("analysis", "data", "derived_data", "genoprobs.rds"))
  saveRDS(dataset.islet.rnaseq$covar, here("analysis", "data", "derived_data", "covar.rds"))
}
```

```{r}
lod_peaks <- readRDS(here("analysis", "data", "derived_data", "lod_peaks.rds"))
annots <- readRDS(here("analysis", "data", "derived_data", "annots.rds"))
expr <- readRDS(here("analysis", "data", "derived_data", "expr.rds"))
```

@@

@@

An unanticipated observation is that some points have mediation LOD differences of nearly 10 
and pleiotropy test statistics of nearly 10. With such large pleiotropy test statistics, one 
has reasonable evidence each point representing two separate QTL. It seems biologically 
unlikely that a nonlocal trait and a mediating local trait would not share a pleiotropic QTL.

In looking closer at these six points, we find that five of the six have *Tmem189* as the 
local trait. This raises questions about the *Tmem189* trait. One known issue with the pleiotropy test is its sensitivity to start and end points for the two-dimensional scan. If the scan interval is too small, the test can return misleading results. If *Tmem189* were to have a very broad univariate peak, such that the peak extended beyond the scan interval, then we might observe misleading results in tests involving *Tmem189*.

```{r scan1-Chr2, eval = FALSE}
# read oddball points from interactive plot
tr <- c("Tmem189",
        "Myo15b",
        "Sephs2",
        "Bcmo1",
        "Tmprss4",
        "Cacnb3",
        "Gm12929",
        "Eya2"
        )
tr_tib <- hot_annot %>%
  filter(symbol %in% tr) %>%
  arrange(annot.id)
K <- readRDS(here("analysis", "data", "derived_data", "kinship.rds"))
map <- readRDS(here("analysis", "data", "derived_data", "map.rds"))
covar <- readRDS(here("analysis", "data", "derived_data", "covar.rds"))
aprobs <- readRDS(here("analysis", "data", "derived_data", "genoprobs.rds"))
expr_sm <- expr[ , colnames(expr) %in% tr_tib$annot.id, drop = FALSE]
s1 <- qtl2::scan1(genoprobs = aprobs,
            pheno = expr_sm,
            kinship = K,
            addcovar = covar, 
            cores = 0
)
# zoom in on Chr2 after 150 Mb
# get length of map$`1`
purrr::map(.x = 1:8, .f = function(x) qtl2::plot_scan1(s1, map, lodcolumn = x, chr = 2, main = paste0(tr_tib[x, ])))
qtl2::find_peaks(scan1_output = s1, map = map, threshold = 7) %>% 
  as_tibble() %>%
  filter(chr == 2) %>%
  left_join(hot_annot, by = c("lodcolumn" = "annot.id"))
```

@@

The Chr 5 scatter plot of mediation LOD difference values against pleiotropy test statistics 
reveals a block-diagonal pattern of blue coloring among the nonlocal traits. We observe the 
set of nonlocal traits clustered into two blocks. Within each block, any two traits share a 
QTL, ie, are pleiotropic. Yet, the off-diagonal cells, ie, those nonlocal trait pairs that 
include one trait from each of the two blocks, tend to have greater pleiotropy test 
statistics. Still, not all off-diagonal blocks have large pleiotropy test statistics. The 
mostly dark blue matrix on the diagonal (from column 22 to 99) features no cells with pleiotropy test statistics above 3.2. While we didn't determine p-values for these analyses, prior analyses suggest that pleiotropy test statistics below 3.2 are consistent with p-values above 0.05, and, thus, are consistent with a single pleiotropic QTL.

@@

@@

@@

@@

>>>>

Simple hotspots: Chr 2, Chr 13

Complex hotspots: Chr 5

Undetermined: Chr 7, Chr 11

@@

```{r saveRDS}
if (!file.exists(here("analysis", "data", "derived_data", "expr.rds"))){
  download.file("https://datadryad.org/bitstream/handle/10255/dryad.166774/Attie_DO378_eQTL_viewer_v1.Rdata?sequence=2",
              destfile = here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata")
              )
  load(here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata"))
  saveRDS(dataset.islet.rnaseq$lod.peaks, here("analysis", "data", "derived_data", "lod_peaks.rds"))
  saveRDS(dataset.islet.rnaseq$annots, here("analysis", "data", "derived_data", "annots.rds"))
  saveRDS(dataset.islet.rnaseq$expr, here("analysis", "data", "derived_data", "expr.rds"))
  saveRDS(K, here("analysis", "data", "derived_data", "kinship.rds"))
  saveRDS(map, here("analysis", "data", "derived_data", "map.rds"))
  saveRDS(genoprobs, here("analysis", "data", "derived_data", "genoprobs.rds"))
  saveRDS(dataset.islet.rnaseq$covar, here("analysis", "data", "derived_data", "covar.rds"))
}
```

```{r readRDS}
lod_peaks <- readRDS(here("analysis", "data", "derived_data", "lod_peaks.rds"))
annots <- readRDS(here("analysis", "data", "derived_data", "annots.rds"))
expr <- readRDS(here("analysis", "data", "derived_data", "expr.rds"))
```

@@

### Chr 2 scatter plots

@@

An unanticipated observation is that some points have mediation LOD differences of

### Chr 2 heatmap

### Chr 2 index v position plot

@@

### Chr 5 scatter plots

### Chr 5 heatmap

The Chr 5 heatmap of pleiotropy test statistics 
reveals a block-diagonal pattern of blue coloring among the nonlocal traits. We observe the 
set of nonlocal traits clustered into two blocks. Within each block, any two traits share a 
QTL, ie, are pleiotropic. Yet, the off-diagonal cells, ie, those nonlocal trait pairs that 
include one trait from each of the two blocks, tend to have greater pleiotropy test 
statistics. Still, not all off-diagonal blocks have large pleiotropy test statistics. The 
mostly dark blue matrix on the diagonal (from column 22 to 99) features no cells with pleiotropy test statistics above 3.2. While we didn't determine p-values for these analyses, prior analyses suggest that pleiotropy test statistics below 3.2 are consistent with p-values above 0.05, and, thus, are consistent with a single pleiotropic QTL.

@@

### Chr 5 index v position plot

@@

## Chr 7 results

### Chr 7 scatter plots

### Chr 7 heatmap

### Chr 7 index v position plot

@@

## Chr 11 results

### Chr 11 scatter plots

### Chr 11 heatmap

### Chr 11 index v position plot

@@

## Chr 13 results

### Chr 13 scatter plots

### Chr 13 heatmap

### Chr 13 index v position plot
